---
title: "Accessibility Case Study: Boston Landing"
author: "Carole Turley Voulgaris, Transportation Economics and Finance (SES 5437), Harvard GSD"
csl: https://www.zotero.org/styles/chicago-fullnote-bibliography
suppress-bibliography: yes
output: tint::tintPdf
bibliography: refs.bib
---

```{r setup, include=FALSE}
options(java.parameters = '-Xmx3G')

library(tidyverse)
library(here)
library(tidytransit)
library(sf)
library(r5r)
library(sfheaders)
library(tigris)
library(knitr)

# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
#options(htmltools.dir.version = FALSE)
```

Boston Landing^[This information in this case study is for teaching purposes and I do not guarantee its accuracy.] is a \$500 million-dollar development by NB Development
group for New Balance Athletics. It comprises six properties located
at 40-180 Guest Street in Boston, Massachusetts. These are listed in Table 1.

```{r, echo=FALSE}
tibble(Address = c("40 Guest St",
                   "60 Guest St",
                   "80 Guest St",
                   "100 Guest St",
                   "125 Guest St",
                   "180 Guest St"),
       `Land use` = c("Boston Celtics Practice Facility and Lab Space",
                    "Office/Lab Building",
                    "Warrior Ice Arena and Office/Lab Building",
                    "New Balance Athletics Inc. World Headquarters",
                    "Lantera Residential",
                    "Hotel"),
       `Date opened` = c("2017",
                           "Future",
                           "2016",
                           "2013",
                           "2018",
                           "Future")) |>
  kable(caption = "Properties included in the Boston Landing development")

```


In connection with the development, New Balance Athletics also 
funded the planning and construction of the \$20 million Boston Landing commuter rail station located adjacent to the site. The station opened in 2017.

Figure 1 shows the locations of the properties and the commuter rail station.

```{r make-site-map, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
boundaries2023 <- here("1-accessibility-boston-landing",
                       "data",
                       "site-parcels.geojson")|>
  st_read() 

station <- here("1-accessibility-boston-landing",
                "data",
                "BL station bound.kml") |>
  st_read() |>
  rename(MAP_PAR_ID = Name) |>
  select(MAP_PAR_ID) |>
  st_zm()

site <- rbind(boundaries2023, station) |>
  mutate(label = c("180",
                   "60",
                   "40",
                   "100",
                   "80",
                   "125",
                   "Commuter Rail Station"))

labels <- st_centroid(site)
```

```{r site-map, fig.margin = FALSE, fig.cap = "Boston Landing development and adjacent commuter rail station. Basemap by CartoDB — Map data © OpenStreetMap contributors", fig.width=6, fig.height=2.3, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(site) +
  annotation_map_tile(type ="cartolight", zoomin = -1) +
  geom_sf() + 
  geom_sf_text(data = labels, 
               angle = -10,
               aes(label = label)) +
  annotation_scale(style = "ticks", location = "br",
                     text_cex = 1) +
  annotation_north_arrow(style = north_arrow_minimal) +
  theme_map()
```

# Transitshed area

The development of the Boston Landing commuter rail station substantially
expanded the size of the transitshed for the site during the weekday 
commute periods. Prior to the station opening, the area of the transitshed
at 6pm on a weekday was about 6.3 square kilometers (Figure 2). 
After the station opened, the area of the transitshed at that time 
was about 14.1 square kilometers (Figure 3).

```{r read-before-iso, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}

iso_before <- here("1-accessibility-boston-landing",
                   "data",
                   "bl-iso-before.geojson") |>
  st_read()

iso_after <- here("1-accessibility-boston-landing",
                   "data",
                   "bl-iso-after.geojson") |>
  st_read()

bbox_before <- st_bbox(iso_before)
bbox_after <- st_bbox(iso_after)

area_km2_after = units::set_units(st_area(iso_after), km^2)
area_km2_before = units::set_units(st_area(iso_before), km^2)

routes <- here("1-accessibility-boston-landing",
               "data",
               "bl-transit-routes.geojson") |>
  st_read()
```

```{r before-iso, fig.margin = FALSE, fig.cap = "30-minute transitshed from development site before the Boston landing Station opening", fig.width=6, fig.height=2.6, cache=TRUE, echo=FALSE}
  ggplot(routes) +
    geom_sf(color = "gray", 
            linewidth = 0.5) +
    geom_sf(data = iso_before,
            alpha = 0.5,
            color = NA,
            fill = "pink") +
    geom_sf(data = site,
            fill = "black",
            color = NA) +
    annotation_scale(style = "ticks", location = "br",
                     text_cex = 1) +
        coord_sf(xlim = c(bbox_after["xmin"],
                          bbox_after["xmax"]),
                 ylim = c(bbox_after["ymin"],
                          bbox_after["ymax"])) +
    annotation_north_arrow(style = north_arrow_minimal,
                           location = "tl") +
    theme_map() 

```



```{r after-iso, fig.margin = FALSE, fig.cap = "30-minute transitshed from development site after the Boston landing Station opening", fig.width=6, fig.height=2.6, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

  ggplot(routes) +
    geom_sf(color = "gray", 
            linewidth = 0.5) +
    geom_sf(data = iso_after,
            alpha = 0.5,
            color = NA,
            fill = "pink") +
      geom_sf(data = site,
            fill = "black",
            color = NA) +
          coord_sf(xlim = c(bbox_after["xmin"],
                      bbox_after["xmax"]),
             ylim = c(bbox_after["ymin"],
                      bbox_after["ymax"])) +
      annotation_scale(style = "ticks", location = "br",
                     text_cex = 1) +
  annotation_north_arrow(style = north_arrow_minimal,
                           location = "tl") +
    theme_map() 

```


# Accessibility of worker residences

Commercial developments benefit from being accessible from the residences of 
(potential) workers (and customers). 

The number of workers^[Employees at _any_ U.S. firm, based on employer tax filings]
living within a thirty-minute transit commute of the site increased between 2012
and 2020 for three reasons:

1. The development itself includes residential space;
2. The development included a commuter rail station that expanded the transit
shed to include existing residential developments that were not previously 
accessible by transit;
3. New residential developments opened in the surrounding area, which may have been
prompted in part by proximity to the development and/or the associated commuter rail 
station.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
blocks_any <- here("1-accessibility-boston-landing",
                   "Data",
                   "blocks.geojson") |>
  st_read()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tibble(Area = c("Site",
                "Pre-development transitshed",
                "Post-development transitshed"),
       `2012` = c(
         sum(blocks_any$w2012[blocks_any$site]),
         sum(blocks_any$w2012[blocks_any$before]),
         sum(blocks_any$w2012[blocks_any$after])),
       `2020` = c(
         sum(blocks_any$w2020[blocks_any$site]),
         sum(blocks_any$w2020[blocks_any$before]),
         sum(blocks_any$w2020[blocks_any$after]))) |>
  kable(caption = "Changes in numbers of resident workers from 2012 to 2020",
        format.args = c(big.mark = " "))
```

# Accessibility of employment

Commercial developments benefit from proximity to employment centers (for the
benefits of agglomeration economies, for example) and residential developments
benefit from proximity to employment.

The number of jobs within a thirty-minute transit commute of the site increased between 2012
and 2020 for three reasons:

1. The development itself includes commercial space;
2. The development included a commuter rail station that expanded the transit
shed to include existing jobs that were not previously 
accessible by transit;
3. New jobs were created or relocated to the surrounding area, which may have been
prompted in part by proximity to the development and/or the associated commuter rail 
station.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tibble(Area = c("Site",
                "Pre-development transitshed",
                "Post-development transitshed"),
       `2012` = c(
         sum(blocks_any$e2012[blocks_any$site]),
         sum(blocks_any$e2012[blocks_any$before]),
         sum(blocks_any$e2012[blocks_any$after])),
       `2020` = c(
         sum(blocks_any$e2020[blocks_any$site]),
         sum(blocks_any$e2020[blocks_any$before]),
         sum(blocks_any$e2020[blocks_any$after]))) |>
  kable(caption = "Changes in numbers of jobs from 2012 to 2020",
        format.args = c(big.mark = " "))
```


# Assessed property values

Table 4 shows
the total assessed land and building values for the site in 2012 and 2020.

```{r, echo=FALSE}
tibble(`-` = c("Land value",
                 "Building value",
                 "Total assessed value"),
       `2012` = c("$8 416 500",
                  "$10 303 400",
                  "$18 719 900"),
       `2020` = c("$88 879 900",
                  "$303 617 600",
                  "$392 497 500")) |>
  kable(caption = "Total assessed property values for site parcels",
        align = c("l", "r", "r"))
```


# Questions for discussion

1. Between 2012 and 2020, how much did the site's accessibility to workers increase?
2. How much of that increase can be attributed to each of the following sources?
    a. Residential development on-site?
    b. Off-site residential development?
    c. Expansion of the transitshed?
3. Over that same time period, how much did site's accessibility to jobs increase?
4. How much of that increase can be attributed to each of the following sources?
    a. Commercial development on-site?
    b. Off-site job creation/relocation?
    c. Expansion of the transitshed?
5. Approximately how much of the increase in the site's value
might you attribute to increase in accessibility over that time period? 
6. To what extent do you agree with the claim that accessibility (by all modes, to all possible destination types) is the primary driver of land
value? What other factors could contribute to land value?