---
title: "Fare Equity Case Study: BART and MARTA"
author: "Carole Turley Voulgaris, Transportation Economics and Finance (SES 5437), Harvard GSD"
csl: https://www.zotero.org/styles/chicago-fullnote-bibliography
suppress-bibliography: yes
output: tint::tintPdf
bibliography: refs.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(sf)
library(tigris)

# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
#options(htmltools.dir.version = FALSE)
```

The Bay Area Rapid Transit Sytem (BART) and Metropolitan Atlanta Regional Transit 
Authority (MARTA) were two of the earliest heavy rail transit systems built in 
the modern era of publicly-funded, federally-subsidized urban rail transit
construction. Both systems initially opened for service in the 1970s. They are
also both named with acronyms that sound like a person's name.

Mallett has developed an accounting method to allocate the total cost of 
BART and MARTA operations among links and stations within each system.^[Mallett, Z., 2023. Spatial and Temporal Variability of Rail Transit Costs and Cost Effectiveness. Transportation Research Record, 2677(1), pp.1444-1460.] He has also estimated the 
average fares paid by passengers riding on each link and by passengers boarding and 
alighting at each station.^[Mallett, Z., 2023. Inequitable Inefficiency: A Case Study of Rail Transit Fare Policies. Working Paper.]

# BART system characteristics

The BART rail system comprises 131.4 miles of track serving 50 stations in 
the San Francisco Bay Area of California.

BART charges a distance-based fare structure that charges riders a base fare for
trips trips up to six miles long. For trips longer than six miles, there is an 
additional fare per mile after the first six miles, and lower fare per mile
after the first fourteen miles. 

Figure 1 shows the network layout with two specific stations and links highlighted.

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
tracts <- tidycensus::get_acs(year = 2019,
                              geography = "tract",
                              state = "CA",
                              county = c("San Francisco",
                                                     "Marin",
                                                     "Alameda",
                                                     "San Mateo",
                                                     "Contra Costa",
                                                     "Solano",
                                                     "Napa",
                                                     "Sonoma"),
                              variables = "B06011_001",
                              output = "wide",
                              geometry = TRUE)

BART_links_start <- here("2-rail-fares",
     "data",
     "LinkAttributes_BART.csv") |>
  read_csv() |>
  st_as_sf(coords = c("long_start", "lat_start")) |>
  select(-long_end, -lat_end) |>
  filter(link_start != "C88" & link_end != "C88") 

BART_links_end <- here("2-rail-fares",
     "data",
     "LinkAttributes_BART.csv") |>
  read_csv() |>
  st_as_sf(coords = c("long_end", "lat_end"))  |>
  select(-long_start, -lat_start) |>
  filter(link_start != "C88" & link_end != "C88")

BART_links <- rbind(BART_links_start, BART_links_end) |>
  group_by(link) |>
  summarise(length = mean(length),
            passengers = mean(passengers),
            cost_total = mean(cost_total),
            costpm = mean(costpm),
            costppx = mean(costppx),
            costppm = mean(costppm),
            percentpaid = mean(percentpaid_effective)) |>
  st_cast("LINESTRING") |>
  mutate(cases = case_when(percentpaid == min(percentpaid) ~ 
                             "Warm Springs to Fremont",
                           percentpaid == max(percentpaid) ~ 
                             "Oakland Center to West Oakland",
                           TRUE ~ "Other BART links")) |>
  mutate(cases = factor(cases, levels = c("Warm Springs to Fremont",
                                             "Oakland Center to West Oakland",
                                             "Other BART links"))) |>
  st_set_crs("NAD83")

BART_station <- here("2-rail-fares",
     "data",
     "StationAttributes_BART.csv") |>
  read_csv() |>
  st_as_sf(coords = c("long", "lat"))  |>
  filter(id %in% BART_links_start$link_start |
           id %in% BART_links_start$link_end) |>
    mutate(cases = case_when(percentpaid_effective == 
                               min(percentpaid_effective) ~ paste0(name, "\nStation"),
                           percentpaid_effective == max(percentpaid_effective) ~ 
                             paste0(name, "\nStation"),
                           TRUE ~ ""))  |>
  st_set_crs("NAD83")

case_stations = BART_station |>
  filter(cases != "") |>
    mutate(`Average fare` = paste0("$", formatC(costppx * percentpaid_effective, digits=2, format="f")),
         `Annual cost` = paste0("$", prettyNum(cost_total, big.mark = " ")),
         `Annual ridership` = prettyNum(passengers, big.mark = " ")) |>
  rename(Station = cases) |>
  st_join(tracts) |>
  group_by(Station,
         `Annual ridership`,
         `Average fare`,
         `Annual cost`) |>
  summarise(`Median income of proximate residents` = paste0("$", prettyNum(round(mean(B06011_001E)), big.mark = " ")))

case_links = BART_links |>
  filter(cases != "Other BART links") |>
  mutate(`Average fare` = paste0("$", formatC(costppx * percentpaid, digits=2, format="f")),
         `Annual cost` = paste0("$", prettyNum(cost_total, big.mark = " ")),
         `Annual ridership` = prettyNum(passengers, big.mark = " ")) |>
  rename(Link = cases,
         `Length (miles)` = length) |>
  st_join(tracts) |>
  group_by(Link,
         `Length (miles)`,
         `Annual ridership`,
         `Average fare`,
         `Annual cost`) |>
  summarise(`Median income of proximate residents` = paste0("$", prettyNum(round(mean(B06011_001E)), big.mark = " ")))

bart_bbox = st_bbox(BART_station)

water <- tigris::area_water(state = "CA", county = c("San Francisco",
                                                     "Marin",
                                                     "Alameda",
                                                     "San Mateo",
                                                     "Contra Costa",
                                                     "Solano",
                                                     "Napa",
                                                     "Sonoma"))



```


```{r, echo=FALSE, warning=FALSE, fig.fullwidth = TRUE, fig.cap="BART system map with locations of links and stations of interest"}
library(ggthemes)

ggplot(BART_links) +
  geom_sf(data = water,
          color = NA,
          fill = "lightgray") +
  geom_sf(data = BART_station,
          size = 1) +
  geom_sf(aes(color = cases),
          linewidth = 1) +
  geom_sf_text(data = case_stations,
                aes(label = Station),
               hjust = 1,
               vjust = c(0, 1),
               nudge_x = -0.02,
               nudge_y = 0.01,
               size = 2,
               color = "#E41A1C") +
  geom_sf(data = case_stations,
          color = "#E41A1C",
          size = 1.5) +
  theme_map() +
  coord_sf(xlim = c(as.numeric(bart_bbox["xmin"]) - 0.03, 
                    as.numeric(bart_bbox["xmax"]) + 0.4),
           ylim = c(as.numeric(bart_bbox["ymin"] - 0.01),
                    as.numeric(bart_bbox["ymax"]))) +
  scale_color_manual(values = c("#4DAF4A",
                               "#377EB8",
                               "gray30")) +
  theme(legend.key.height = unit(0.15, "in"),
        legend.key.width = unit(0.3, "in"),
        legend.title = element_blank(),
        legend.background = element_blank(),
        legend.key = element_rect(fill = NA),
        legend.position = c(0.6, 0.75),
        legend.text = element_text(size = 5))
```

Table 1 shows some basic data on the lines highlighted in Figure 1 and Table 2
shows some basic data on the stations.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
case_links |>
  st_drop_geometry() |>
  `rownames<-`(case_links$Link) |>
  select(-Link) |>
  t() |>
  kable(align = c('r', 'r', 'l'),
        caption = "Characteristics of BART links of interest") |>
  kable_styling(full_width = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
case_stations |>
  st_drop_geometry() |>
  `rownames<-`(case_links$Station) |>
  select(-Station) |>
  t() |>
  kable(align = c('r', 'r', 'l'),
        caption = "Characteristics of BART stations of interest") |>
  kable_styling(full_width = TRUE)
```

# MARTA system characteristics

The MARTA rail system comprises 48 miles of track serving 38 stations in 
the Atlanta, Georgia region.

MARTA charges a flat fare of $2.50 per trip, although the average fare on the system
overall and for riders on a particular part of the system is generally lower that that
due to fare discounts and the use of multi-day transit passes. 

Figure 1 shows the network layout with two specific stations and links highlighted.

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
tracts <- tidycensus::get_acs(year = 2019,
                              geography = "tract",
                              state = "GA",
                              county = c("Fulton", "DeKalb"),
                              variables = "B06011_001",
                              output = "wide",
                              geometry = TRUE)

MARTA_links_start <- here("2-rail-fares",
     "data",
     "LinkAttributes_MARTA.csv") |>
  read_csv() |>
  st_as_sf(coords = c("long_start", "lat_start")) |>
  select(-long_end, -lat_end)  

MARTA_links_end <- here("2-rail-fares",
     "data",
     "LinkAttributes_MARTA.csv") |>
  read_csv() |>
  st_as_sf(coords = c("long_end", "lat_end"))  |>
  select(-long_start, -lat_start) 

MARTA_links <- rbind(MARTA_links_start, MARTA_links_end) |>
  group_by(link) |>
  summarise(length = mean(length),
            passengers = mean(passengers),
            cost_total = mean(cost_total),
            costpm = mean(costpm),
            costppx = mean(costppx),
            costppm = mean(costppm),
            percentpaid = mean(percentpaid_effective)) |>
  st_cast("LINESTRING") |>
  mutate(cases = case_when(link == "F10-F11" ~ 
                             "North Springs to Sandy Springs",
                           link == "N3-N4" ~ 
                             "North Avenue to Midtown",
                           TRUE ~ "Other MARTA links")) |>
  mutate(cases = factor(cases, levels = c("North Springs to Sandy Springs",
                                             "North Avenue to Midtown",
                                             "Other MARTA links"))) |>
  st_set_crs("NAD83")

MARTA_station <- here("2-rail-fares",
     "data",
     "StationAttributes_MARTA.csv") |>
  read_csv() |>
  st_as_sf(coords = c("long", "lat"))  |>
  filter(id %in% MARTA_links_start$link_start |
           id %in% MARTA_links_start$link_end) |>
    mutate(cases = case_when(name == "North Springs" ~ paste0(name, " Station"),
                           percentpaid_effective == max(percentpaid_effective) ~ 
                             paste0(name, " Station"),
                           TRUE ~ ""))  |>
  st_set_crs("NAD83")

case_stations = MARTA_station |>
  filter(cases != "") |>
    mutate(`Average fare` = paste0("$", formatC(costppx * percentpaid_effective, digits=2, format="f")),
         `Annual cost` = paste0("$", prettyNum(cost_total, big.mark = " ")),
         `Annual ridership` = prettyNum(passengers, big.mark = " ")) |>
  rename(Station = cases) |>
  st_join(tracts) |>
  group_by(Station,
         `Annual ridership`,
         `Average fare`,
         `Annual cost`) |>
  summarise(`Median income of proximate residents` = paste0("$", prettyNum(round(mean(B06011_001E)), big.mark = " ")))

case_links = MARTA_links |>
  filter(cases != "Other MARTA links") |>
  mutate(`Average fare` = paste0("$", formatC(costppx * percentpaid, digits=2, format="f")),
         `Annual cost` = paste0("$", prettyNum(cost_total, big.mark = " ")),
         `Annual ridership` = prettyNum(passengers, big.mark = " "),
         `Length (miles)` = round(length, digits = 2)) |>
  rename(Link = cases) |>
  st_join(tracts) |>
  group_by(Link,
         `Length (miles)`,
         `Annual ridership`,
         `Average fare`,
         `Annual cost`) |>
  summarise(`Median income of proximate residents` = paste0("$", prettyNum(round(mean(B06011_001E)), big.mark = " ")))

marta_bbox = st_bbox(MARTA_station)

water <- tigris::area_water(state = "GA", county = c("Fulton", "DeKalb"))



```


```{r, echo=FALSE, warning=FALSE, fig.fullwidth = TRUE, fig.cap="BART system map with locations of links and stations of interest"}
library(ggthemes)

ggplot(MARTA_links) +
  geom_sf(data = water,
          color = NA,
          fill = "lightgray") +
  geom_sf(data = MARTA_station,
          size = 1) +
  geom_sf(aes(color = cases),
          linewidth = 1) +
  geom_sf_text(data = case_stations,
                aes(label = Station),
               hjust = 0,
               vjust = 0.5,
               nudge_x = 0.01,
               nudge_y = 0,
               size = 2,
               color = "#E41A1C") +
  geom_sf(data = case_stations,
          color = "#E41A1C",
          size = 1.5) +
  theme_map() +
  coord_sf(xlim = c(as.numeric(marta_bbox["xmin"]) - 0.03, 
                    as.numeric(marta_bbox["xmax"]) + 0.4),
           ylim = c(as.numeric(marta_bbox["ymin"] - 0.01),
                    as.numeric(marta_bbox["ymax"]))) +
  scale_color_manual(values = c("#4DAF4A",
                               "#377EB8",
                               "gray30")) +
  theme(legend.key.height = unit(0.15, "in"),
        legend.key.width = unit(0.3, "in"),
        legend.title = element_blank(),
        legend.background = element_blank(),
        legend.key = element_rect(fill = NA),
        legend.position = c(0.6, 0.75),
        legend.text = element_text(size = 5))
```

Table 1 shows some basic data on the lines highlighted in Figure 1 and Table 2
shows some basic data on the stations.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
case_links |>
  st_drop_geometry() |>
  `rownames<-`(case_links$Link) |>
  select(-Link) |>
  t() |>
  kable(align = c('r', 'r', 'l'),
        caption = "Characteristics of BART links of interest") |>
  kable_styling(full_width = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
case_stations |>
  st_drop_geometry() |>
  `rownames<-`(case_links$Station) |>
  select(-Station) |>
  t() |>
  kable(align = c('r', 'r', 'l'),
        caption = "Characteristics of BART stations of interest") |>
  kable_styling(full_width = TRUE)
```




# Questions for discussion

1. For each of the highlighted links on the MARTA system:
    a. What is the average cost per passenger-mile?
    b. What is the average fare collected per passenger-mile?
    c. What is the average subsidy per passenger-mile?
    d. Assuming a person living near each link makes 500 trips on the link each year, 
    what is the average subsidy the person living near each link receives in absolute 
    dollars and as a percentage of the median income for the area where they live?
    
2. For each of the highlighted links on the BART system:
    a. What is the average cost per passenger-mile?
    b. What is the average fare collected per passenger-mile?
    c. What is the average subsidy per passenger-mile?
    d. Assuming a person living near each link makes 500 trips on the link each year, 
    what is the average subsidy the person living near each link receives in absolute 
    dollars and as a percentage of the median income for the area where they live?

3. What additional data would you want in order to more fully evaluate the equity 
of transit subsidies in these two regions?