---
title: "Establishment of new informal transport service"
author: "Carole Turley Voulgaris, Transportation Economics and Finance (SES 5437), Harvard GSD"
csl: https://www.zotero.org/styles/chicago-fullnote-bibliography
suppress-bibliography: yes
output: tint::tintPdf
bibliography: refs.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(sf)
library(tigris)

# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
#options(htmltools.dir.version = FALSE)
```

# Background

This case is based on research conducted by Tamara Kerzhner.

To test a theory that there may be a market for minibus routes that aren't 
currently being served, researchers paid operators a subsidy to operate a new
radial route in Kampala, Uganda.

Operators were paid 211,000 shillings (about \$56 USD) per 
day to operate the new route during the first half of a 
three-week pilot, and 150,000 shillings (about \$40 USD)
per day during the second half of the pilot. After the pilot concluded, 
some drivers continued to operate the route without a subsidy.

# Operating data

The following table summarizes the following operating data for each day during the 
pilot.

* Day: The day of the pilot program (a value of 1 is the first day of the pilot)
* Day of week: The day of the week
* Vehicles operating: The number of vehicles serving the route on that day
* Runs per day: The total number of runs (a vehicle's trip from one end of th
route to the other in either direction) by all vehicles that day
* Passengers: The total number of passengers served by all vehicles that
day, based on data reported by the vehicle operators (operators probably
misrepresented this data to be lower than the actual number of passengers)
* Fare collected: The total fare collected by all vehicles on that day, 
based on data reported by the vehicle operators (operators probably underreported
these values)
* Subsidy payment: The total payment by researchers to all vehicle operators 
to cover vehicle operating costs.

```{r, echo=FALSE, message=FALSE, fig.fullwidth=TRUE}
data <- here("Informal",
             "vehicles_anon.csv") |>
  read_csv() |>
  mutate(revenue = 500*`500`+
                   1000*`1000` +
                   1500*`1500` +
                   2000*`2000` +
                   2500*`2500` +
                   3000*`3000` +
                   4000*`4000`) |>
  mutate(rev_per_run = revenue / Runs) 

veh_by_day <- data |>
  group_by(Day) |>
  summarise(`Vehicles operating` = n(),
            `Runs per day` = sum(Runs),
            `Passengers` = sum(Women) + sum(Men) + sum(Child),
            `Fare collected` = sum(revenue)) |>
  mutate(`Day of week` = c(rep(c("Monday", 
                                     "Tuesday",
                                     "Wednesday",
                                     "Thursday",
                                     "Friday",
                                     "Saturday",
                                     "Sunday"), 3), "Monday")) |>
  mutate(sub_per_veh = c(rep(211000, 10), rep(150000, 12))) |>
  mutate(`Subsidy payment` = sub_per_veh * `Vehicles operating`) |>
  select(`Day`,
         `Day of week`,
         `Vehicles operating`,
         `Runs per day`,
         `Passengers`,
         `Fare collected`,
         `Subsidy payment`)

kable(veh_by_day) |>
  kable_styling(full_width = TRUE)
```

After the pilot program (and associated subsidies) ended, the route 
continued to operate with 5 vehicles per day on week days and market
days and 3 vehicles per day on all other days.

# Questions for discussion

1. What was the average daily revenue per run (including
both fares and subsidies) during the first week of the study?

2. What was the average daily revenue per run (including
both fares and subsidies) during the last week of the study?

3. What was the average number of daily runs during the first week 
of the study?

4. What was the average number of daily runs during the last week 
of the study?

```{r}
change <- veh_by_day |>
  mutate(week = c(rep(1, 7), rep(2, 8), rep(3, 7))) |>
  filter(week == 1 | week == 3) |>
  mutate(rev_per_run = (`Fare collected` + `Subsidy payment`)/ `Runs per day`) |>
  group_by(week) |>
  summarise(`Average revenue per run` = mean(rev_per_run),
            `Average daily runs` = mean(`Runs per day`)) 

kable(change)


```

5. Based on the data presented here, what would be a reasonable 
estimate for total number of daily runs that continued in a typical
week after the conclusion of the study?

_Assuming one weekend day is a market day? 33 vehicles per week, each making 4 runs per day? So 132 runs per week, or 18.9 runs per day?_

6. Assuming a constant price elasticity of supply, what is a 
reasonably guess for the fare revenue that the route generated
after the pilot concluded?

Percent change in price:

```{r}
pct_price <- ((change$`Average revenue per run`[1] - 
                change$`Average revenue per run`[2]) * 2) /
  (change$`Average revenue per run`[1] + 
                change$`Average revenue per run`[2])
```

Percent change in supply:

```{r}
pct_supply <- ((change$`Average daily runs`[1] - 
                change$`Average daily runs`[2]) * 2) /
  (change$`Average daily runs`[1] + 
                change$`Average daily runs`[2])
```

Elasticity:

```{r}
elast = pct_supply / pct_price

elast
```

A 28% reduction in revenue per trip resulted in a 51% reduction in 
supply. This is very elastic supply.

Increase in supply after pilot:

```{r}
inc_supply <- ((132/7 - 
                change$`Average daily runs`[2]) * 2) /
  (132/7 + 
                change$`Average daily runs`[2])
```

Likely increase in revenue per trip:

```{r}
inc_rev = inc_supply / elast

inc_rev
```

Revenue per trip probably increased by 3.5%

```{r}
new_rev_per_trip = change$`Average revenue per run`[2]*(1+inc_rev)

new_rev = new_rev_per_trip*(132/7)
```

None of this makes much sense, since there's probably some substantial
underreporting of fares.

7. Given that there appears to have been a market for this route,
how would you explain why operators had not already begun to serve it?

Risk
