---
title: "Elasticity Case Study: Stockholm Congestion Charge"
author: "Carole Turley Voulgaris, Transportation Economics and Finance (SES 5437), Harvard GSD"
csl: https://www.zotero.org/styles/chicago-fullnote-bibliography
suppress-bibliography: yes
output: tint::tintPdf
bibliography: refs.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
#options(htmltools.dir.version = FALSE)
```

In 2006, Stockholm introduced a congestion charging system that charged 
drivers a toll when crossing a cordon line around the inner city. In addition to the material presented in this case study please read
chapter 7 of *Transforming Urban Transport*^[Davis, Diane E., and Alan Altschuler, eds. Transforming Urban Transport. Oxford University Press, USA, 2019.] for further background on the Stockholm congestion charges. Data in this case study are from Börjesson and Kristoffersson (2018).^[Börjesson, Maria, and Ida Kristoffersson. "The Swedish congestion charges: Ten years on." Transportation Research Part A: Policy and Practice 107 (2018): 35-51.]

The level of congestion charge varies by time of day, but remained at
a consistent level from day to day from 2006 to 2015. In 2016, congestion 
charge level increased. Table 1 shows congestion charges by time of 
day before and after the 2016 increase.

```{r, echo=FALSE}
tibble(Time = c("6:30am - 6:59am",
                   "7:00am - 7:29am",
                   "7:30am - 8:29am",
                   "8:30am - 8:59am",
                   "9:00am - 9:29am",
                   "9:30am - 2:59pm",
                   "3:00pm - 3:29pm",
                   "3:30pm - 3:59pm",
                   "4:00pm - 5:29pm",
                   "5:30pm - 5:59pm",
                   "6:00pm - 6:29pm",
                   "6:30pm - 6:29am"),
       `Cordon charge, 2006 - 2016` = c("SEK 10",
                    "SEK 15",
                    "SEK 20",
                    "SEK 15",
                    "SEK 10",
                    "SEK 10",
                    "SEK 10",
                    "SEK 15",
                    "SEK 20",
                    "SEK 15",
                    "SEK 10",
                    "SEK 0"),
       `Cordon charge, 2016 - present` = c("SEK 15",
                    "SEK 25",
                    "SEK 35",
                    "SEK 25",
                    "SEK 15",
                    "SEK 11",
                    "SEK 15",
                    "SEK 25",
                    "SEK 35",
                    "SEK 25",
                    "SEK 15",
                    "SEK 0")) |>
  kable(caption = "Stockholm congestion charges by year and time of day",
        align = c("l", "c", "c")) |>
  kable_styling("striped")

```

Based on average trip lengths and vehicle operating cost, the cost of a typical
trip across the cordon line, not including the congestion charge, would
have been SEK 25.5 in 2005/2006 and SEK 31.5 in 2015/2016. 

In Stockholm, about 10 percent of households have access to a "company car":
a car that is owned by a household member's employer and that the household
has access to as a perk of employment. When a driver crosses the cordon
in a company car, their employer pays the charge. Alternative-fuel vehicles
were exempt from paying the charge until 2012 and taxis were exempt from the charge until 2007.


Table 2 shows the volume of non-exempt vehicles crossing the toll cordon 
during in 2005 (immediately before the congestion charge was implemented) and
2006 (immediately after the charge was implemented).The 
average toll paid by non-exempt vehicles in 2006 was SEK 12.8.

```{r, echo=FALSE}
tibble(Year = c("2005",
                   "2006"),
       `Average vehicle volume (veh/hr)` = 
         c("30,021",
                    "21,114")) |>
  kable(caption = "Traffic volumes across the cordon before and after congestion charge implementation",
        align = c("l", "c"))

```

Table 3 shows the peak-period traffic volume of private vehicles across
the cordon in 2015 (immediately before the charge
increased), and 2016 (immediately after the charge increased). It also shows the average toll paid by private vehicles in the peak period
for each year.

```{r, echo=FALSE}
tibble(Year = c("2015",
                   "2016"),
       `Average vehicle volume during peak (veh/hr)` = 
         c("13,570",
                    "11,878"),
       `Average cordon charge` = c("SEK 17.9",
                    "SEK 30.7")) |>
  kable(caption = "Average volumes and charges for ",
        align = c("l", "c", "c"))

```

# Questions for discussion

1. By what percentage did the cost of traveling across the inner-city cordon
change:
    a. For all trips in non-exempt vehicles between 2005 and 2006? 
    b. For peak-period trips in private vehicles between 2015 and 2016?
2. By what percent did the volume trips crossing the 
cordon change:
    a. For all trips in non-exempt vehicles between 2005 and 2006? 
    b. For peak-period trips in private vehicles between 2015 and 2016?
3. What was the price elasticity of demand for travel across the cordon:
    a. For all trips in non-exempt vehicles between 2005 and 2006? 
    b. For peak-period trips in private vehicles between 2015 and 2016?
4. What was the change in consumer surplus associated with:
    a. The introduction of the congestion charge in 2006?
    b. The increase in the congestion charge in 2016?
5. For the consumer surplus changes in both 2006 and 2016,
how much of the change was experienced by people who continued to drive
across the cordon and how much was experienced by those who shifted to 
other modes or reduced their trips to and from the inner city?
6. What other information would you need to fully account for the costs
and benefits of the congestion charge for people who live or work
in Stockholm?

