---
title: "Demand, Elasticity, and Consumer Surplus"
author: "Carole Turley Voulgaris, Transportation Economics and Finance (SES 5437), Harvard GSD"
csl: https://www.zotero.org/styles/chicago-fullnote-bibliography
suppress-bibliography: yes
output: tint::tintPdf
bibliography: refs.bib
---



```{r setup, include=FALSE}
library(tidyverse)
library(here)

# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
#options(htmltools.dir.version = FALSE)
```

When evaluating the effects of changes to the transportation system, it is 
useful to understanding how one might measure the total economic cost of 
travel, to predict how demand changes in 
response to changes in cost, and to quantify how the benefits people get 
from the transportation system change when the cost of travel changes.

# Cost of travel

The total cost of travel includes both *monetary* and *non-monetary* costs.

*Monetary* costs include things like transit fares, tolls, and the cost
of fuel and vehicle maintenance. 

*Non-monetary* costs include things like travel time, discomfort, and risk of
death, injury, or property damage. In many common circumstances, travel time
represents most of the total cost of travel. Risk of death or injury may also
be an important factor.

In order to combine monetary and non-monetary costs, we need to be able to 
express them with a common unit of measurement. One way to do this is to assign
a monetary value to non-monetary costs.

# Demand for travel

We can observe the demand for travel as the total amount of travel at the 
existing price. The units of measurement for travel demand should be consistent
with with the data we have on prices. If we have data on the price per trip, 
then our measure of travel demand should be the number of trips. If we have
data on the price per distance traveled (e.g. the cost per mile or the cost
per kilometer), then our measure of travel demand should be the total distance
traveled.

# Demand curves

If the price of travel changes, it is possible that the demand for 
travel will change as well. 

A demand curve like the one shown in Figure 1 illustrates the relationship 
between demand and price, with 
price shown on the y-axis and demand shown on the x-axis. We typically
plot independent variables on the x-axis and dependent variables on the 
y-axis. If we understand demand changes to be in response to price changes,
this convention of plotting price on the y-axis will seem backwards (or 
sideways).

```{r ex-demand, fig.margin = TRUE, fig.cap = "Example of a demand curve", fig.width=3.5, fig.height=3.5, cache=TRUE, echo=FALSE, message=FALSE}

demand <- tibble(`Cost per trip` = 1*(1.01^(1:100)),
                 `Number of trips taken` = 2000*(0.99^(1:100)))

ggplot(demand) +
  geom_line(aes(x = `Number of trips taken`, y = `Cost per trip`)) +
  scale_y_continuous(breaks = breaks <- seq(1, 2.75, by=0.25),
                     labels = paste0("$", formatC(breaks, digits = 3))) +
  theme_minimal()
```

# Elasticity

The *elasticity* of demand refers to the steepness of the demand curve
or how responsive people are to price changes. If people always use a
particular transportation system the same amount, no matter how much 
it costs, then we would say that demand is *perfectly inelastic* and the demand
curve would be a vertical line like the one in Figure 2, indicating that
demand is the same at all prices.

```{r inelas-demand, fig.margin = TRUE, fig.cap = "Perfectly inelastic demand curve", fig.width=3.5, fig.height=3.5, cache=TRUE, echo=FALSE, message=FALSE}

inelastic_demand <- tibble(`Cost per trip` = c(1, 2.7),
                 `Number of trips taken` = c(1500, 1500))

ggplot(inelastic_demand) +
  geom_line(aes(x = `Number of trips taken`, y = `Cost per trip`)) +
  scale_x_continuous(limits = c(750, 2000),
                     breaks = c(1000, 1500, 2000)) +
  theme_minimal() 

```

Inelastic demand would suggest that there are *no* alternatives to travel 
on a particular system, and this is never the case (staying home is often a 
reasonable alternative). However, when there are very few alternatives,
a demand curve would be very steep. Figure 3 shows a demand curve for gasoline
in the United States between 2004 and 2014^[**Source:** Based on quarterly data from Eliana Eitches and Vera Crain , “Using gasoline data to explain inelasticity ,” Beyond the Numbers: Prices & Spending, vol. 5, no. 5 (U.S. Bureau of Labor Statistics, March 2016), https://www.bls.gov/opub/btn/volume-5/using-gasoline-data-to-explain-inelasticity.htm].

```{r gas-demand, fig.margin = TRUE, fig.cap = "Demand curve for gasoline in the United States", fig.width=3.5, fig.height=3.5, cache=TRUE, echo=FALSE, message=FALSE}

gas_data <- here("2-elasticity-stockholm",
                 "gas-prices.csv") |>
  read_csv() |>
  rename(total_spend = 
           `Dollar expenditures for gasoline and gasoline on out of town trips`,
         gallons = `Estimated gallons of gasoline purchased using CPI data`) |>
  mutate(price_per_gal = total_spend / gallons) |>
  mutate(price_pct_of_max = price_per_gal / max(price_per_gal)) |>
  mutate(gal_pct_of_max = gallons / max(gallons))

ggplot(gas_data, aes(x = gal_pct_of_max, y = price_pct_of_max)) +
  geom_point(size = 0.5) +
  stat_smooth(method = "lm",
              linewidth = 0.1,
              color = "black",
              lty = "dashed") +
  scale_x_continuous(limits = c(0.4, 1),
                     name = "Gallons of gasoline purchased\n(as a percent of the maximum\nvalue between 2004 and 2014)",
                     breaks = breaks <- seq(0.4, 1, by = 0.2),
                     labels = paste0(breaks*100, "%")) +
  scale_y_continuous(limits = c(0.4, 1),
                     name = "Average price per gallon\n(as a percent of the maximum\nvalue between 2004 and 2014)",
                     breaks = breaks <- seq(0.4, 1, by = 0.2),
                     labels = paste0(breaks*100, "%")) +
  theme_minimal()

```

Between 2004 and 2014, gas prices fluctuated between \$3.91 per gallon (in
the third quarter of 2008) and \$1.72 per gallon (in the first quarter of 
2004). During the first quarter of 2004, when gas cost about \$1.72 per gallon,
the average household consumed about 210.7 gallons of gasoline. During the the 
third quarter of 2008, when gas prices were 127 percent higher, the average 
household consumed 180.5 gallons of gasoline: an reduction of just 14 percent.

We can calculate elasticity as the ratio of an observed change in demand to 
the associated observed change in price, or the reduction in demand that would 
be associated with a one-percent increase in price.

In the above example, the elasticity of demand for gasoline would be:

$$
\frac{14\%}{127\%} = 0.11
$$

It's worth noting that, while \$3.91 is 127% higher than \$1.72, 
\$1.72 is 56% lower than \$3.91. Likewise, 180.5 gallons is 14% 
lower than 210.7 gallons, but 210.7 gallons is 17% higher than 
180.5 gallons. This means that if we were talking about a decrease 
in gas prices rather than an increase, we'd get an elasticity of:

$$
\frac{17\%}{56\%} = 0.30
$$

It doesn't really make sense that price elasticities would be 
different for price increases than decreases. A common way 
to bring them into alignment is to calculate the percent 
change as a symmeterical percent change:

$$
change = \frac{difference}{average}
$$

So the symmetrical percent change in gas prices would be:

$$
\frac{3.91 - 1.72}{\frac{1}{2}(3.91+1.75)} = 77\%
$$

The symmetrical percent change in fuel consumption would be:

$$
\frac{210.7 - 180.5}{\frac{1}{2}(210.7+180.5)} = 15\%
$$

And the elasticity (for a price increase or decrease) would be:

$$
\frac{15\%}{77\%} = 0.19
$$

When you are setting prices for something, having a guess as to 
how elastic the demand for that thing is can be really useful. 
When there is inelastic demand, you can expect revenue to increase
by the same amount that the price increases. If demand is elastic,
small increases in price can lead to large increases in demand, 
so changes in revenue might be modest (or even negative).

# Consumer surplus

You can think of the downward sloping demand curve as implying 
that each person has a price they would be willing to pay for a 
given service or good (this is a simplification).

In the demand curve shown in Figure 1, at a price of \$2 per trip, 
there are about 990 people who would be willing to make a trip. At 
a price of a \$1.99 per trip, 1000 people would be willing to make
a trip. These additional ten people who are willing to pay \$1.99
for a trip (but not \$2) are satisfied with a price of \$1.99: They
are paying exactly as much for the trip as the trip is worth to them.

The other 990 people who would have still made a trip at a price of 
\$2 per trip are even happier because they are paying less for the trip 
than they were willing to pay. Some of them would have been willing to 
pay more than \$2.50 per trip, so those folks are very happy with the
price of \$1.99. Others would have been willing to pay up to $2, so 
they happy about the price being \$1.99, but not as happy as those
who would have been willing to pay up to \$1.

The total difference between the actual price of something and how much 
everyone who purchases at that price would have been willing to 
pay is the *consumer surplus*. You can picture it as the area below
the demand curve and above the price level, as shown in Figure 4.

```{r surplus, fig.margin = TRUE, fig.cap = "Illustration of consumer surplus", fig.width=3.5, fig.height=3.5, cache=TRUE, echo=FALSE, message=FALSE}

demand <- demand |>
  mutate(price1 = 1.75,
         price2 = 2)

ggplot(demand) +
  geom_line(aes(x = `Number of trips taken`, y = `Cost per trip`)) +
  geom_line(data = demand[demand$`Number of trips taken` < 1500,],
            aes(x = `Number of trips taken`, y = `price1`)) +
  geom_ribbon(data = demand[demand$`Cost per trip` > demand$price1,],
              aes(ymin=price1,
                  ymax=`Cost per trip`, 
                  x = `Number of trips taken`), 
              fill="gray", alpha=0.5) +
  scale_y_continuous(breaks = breaks <- seq(1, 2.75, by=0.25),
                     labels = paste0("$", formatC(breaks, digits = 3))) +
  annotate("text", x = 1530, y = 1.75, label = "Price = $1.75",
           hjust = 0) +
  theme_minimal()
```

Consumer surplus is most useful for thinking about the total 
benefit (or loss) to consumers that comes with a change in
price. Figure 5 illustrates the increase in consumer surplus that
comes from reducing the price per trip from \$2 per trip to \$1.75
per trip. The dark-gray area is the consumer surplus when the price is \$2 and
the light-gray area is the additional consumer surplus that is gained
when the price drops to \$1.75.

```{r surplus-change, fig.margin = TRUE, fig.cap = "Illustration of a change in consumer surplus", fig.width=3.5, fig.height=3.5, cache=TRUE, echo=FALSE, message=FALSE}

plot <- ggplot(demand) +
  geom_line(aes(x = `Number of trips taken`, y = `Cost per trip`)) +
  geom_line(data = demand[demand$`Number of trips taken` < 1250,],
            aes(x = `Number of trips taken`, y = `price1`)) +
    geom_line(data = demand[demand$`Number of trips taken` < 1250,],
            aes(x = `Number of trips taken`, y = `price2`)) +
  geom_ribbon(data = demand[demand$`Cost per trip` > demand$price1,],
              aes(ymin=price1,
                  ymax=`Cost per trip`, 
                  x = `Number of trips taken`), 
              fill="gray70", alpha=0.5) +
    geom_ribbon(data = demand[demand$`Cost per trip` > demand$price2,],
              aes(ymin=price2,
                  ymax=`Cost per trip`, 
                  x = `Number of trips taken`), 
              fill="gray30", alpha=0.5) +
  scale_y_continuous(breaks = breaks <- seq(1, 2.75, by=0.25),
                     labels = paste0("$", formatC(breaks, digits = 3))) +
  annotate("text", x = 1280, y = 2, label = "Initial price = $2.00",
           hjust = 0) +
  annotate("text", x = 1280, y = 1.75, label = "Final price = $1.75",
           hjust = 0) +
  theme_minimal()

plot
```

We can quantify the benefit of this price change to consumers
by calculating the area of the light-gray area, which we can 
approximate as a rectangle plus a triangle - as illustrated in 
in Figure 6.

```{r surplus-parts, fig.margin = TRUE, fig.cap = "Parts of consumer surplus", fig.width=3.5, fig.height=3.5, cache=TRUE, echo=FALSE, message=FALSE}

plot +
  annotate("rect", 
           xmin = min(demand$`Number of trips taken`),
           xmax = 990,
           ymin = 1.75, ymax = 2,
           fill = NA,
           linetype = "dotted",
           color = "black",
           linewidth = 1.1) +
  annotate("segment",
           x = 1139,
           y = 1.75,
           xend = 990,
           yend = 2,
           linetype = "dotted",
           color = "black",
           linewidth = 1.1) +
  annotate("segment",
           x = 1139,
           y = 1.75,
           xend = 990,
           yend = 1.75,
           linetype = "dotted",
           color = "black",
           linewidth = 1.1)
```

The area of the rectangle is the demand at the initial price times
the change in price. The area of the triangle is one half the
difference in the price times the difference in demand.

Dividing the area into a rectangle and a triangle isn't just 
a trick for calculating the area. It also has a direct interpretation.
The area of the rectangle represents the benefit to existing
travelers, and the area of the triangle represents the benefit to 
new travelers.